Very simplified likelihood

### 2024-03-13 inst/hankin-ejs.tex
tidyup in prep. for submission

\documentclass[ejs]{imsart}

@@

\arxiv{2010.00000}
\startlocaldefs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Uncomment next line to change            %%
%% the type of equation numbering           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\numberwithin{equation}{section}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Axiom, Claim, Corollary, Hypothesis, %%
%% Lemma, Theorem, Proposition              %%
%% use \theoremstyle{plain}                 %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{axiom}{Axiom}
\newtheorem{claim}[axiom]{Claim}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Assumption, Definition, Example,     %%
%% Notation, Property, Remark, Fact         %%
%% use \theoremstyle{definition}            %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem*{example}{Example}
\newtheorem*{fact}{Fact}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Case use \theoremstyle{remark}       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{remark}
\newtheorem{case}{Case}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Please put your definitions here:        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\endlocaldefs

@@

\begin{abstract}
Here a simplification of the Plackett-Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Placket-Luce strength of $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem appears to have a wide range of applications and I
present some analysis of datasets drawn from Formula 1 motor racing
and community initiative Parkrun.
\end{abstract}

\begin{keyword}[class=MSC]
\kwd[Primary ]{6202}
\kwd{00X00}
\kwd[; secondary ]{62F30}
\end{keyword}

\begin{keyword}
\kwd{First keyword}
\kwd{second keyword}
\end{keyword}

@@

(that is, one minus the mean ratio of rank to number of [cloned]
competitors).  Previous related work presented the case $n=2$ [in
  current notation].  Considering the case $n=7$ gives us $7+1=8$
distinct likelihood functions, Figure~\ref{ninelikes}.  We see an
unanticipated asymmetry between $r$ and $n-r$.  For $r=3$ one might
expect that $\hat{a}=\frac{1}{2}$ on the grounds that the focal
competitor's performance was median.  However, $\hat{a}\simeq 0.64$.

\begin{figure}[t]
\includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supplementary Material, including data   %%
%% sets and code, should be provided in     %%
%% {supplement} environment with title      %%
%% and short description. It cannot be      %%
%% available exclusively as external link.  %%
%% All Supplementary Material must be       %%
%% available to the reader on Project       %%
%% Euclid with the published article.       %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{supplement}
\stitle{Title of Supplement A}
\sdescription{Short description of Supplement A.}
\end{supplement}
\begin{supplement}
\stitle{Title of Supplement B}
\sdescription{Short description of Supplement B.}
\end{supplement}

>>>>

\documentclass[ejs,noshowframe]{imsart}

@@

\startlocaldefs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Uncomment next line to change            %%
%% the type of equation numbering           %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\numberwithin{equation}{section}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Axiom, Claim, Corollary, Hypothesis, %%
%% Lemma, Theorem, Proposition              %%
%% use \theoremstyle{plain}                 %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{axiom}{Axiom}
\newtheorem{claim}[axiom]{Claim}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Assumption, Definition, Example,     %%
%% Notation, Property, Remark, Fact         %%
%% use \theoremstyle{definition}            %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem*{example}{Example}
\newtheorem*{fact}{Fact}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% For Case use \theoremstyle{remark}       %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{remark}
\newtheorem{case}{Case}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Please put your definitions here:        %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\endlocaldefs

@@

\begin{abstract}
Here a simplification of the Plackett-Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Placket-Luce strength of $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem appears to have a wide range of applications and I
present some analysis of datasets drawn from Formula~1 motor racing
and community initiative Parkrun.
\end{abstract}

\begin{keyword}[class=MSC]
\kwd[Primary ]{6202}
\kwd{Statistical inference under constraints}
\kwd[ ; secondary: Plackett-Luce likelihood functions; ]{62F30}
\end{keyword}

@@

(that is, one minus the mean ratio of rank to number of [cloned]
competitors).  Previous related work presented the case $n=2$ [in
  current notation].  Considering the case $n=7$ gives us $7+1=8$
distinct likelihood functions, Figure~\ref{ninelikes}.  We see an
unanticipated asymmetry between $r$ and $n-r$.  For $r=4$ one might
expect that $\hat{a}=\frac{1}{2}$ on the grounds that the focal
competitor's performance was median.  However, $\hat{a}\simeq 0.5886$.

\begin{figure}[t]
\includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}.  Value of $r$ indicated in colour by its curve; vertical line shows $\hat{a}\simeq 0.5886$ for the case $r=4$}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Supplementary Material, including data   %%
%% sets and code, should be provided in     %%
%% {supplement} environment with title      %%
%% and short description. It cannot be      %%
%% available exclusively as external link.  %%
%% All Supplementary Material must be       %%
%% available to the reader on Project       %%
%% Euclid with the published article.       %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{supplement}
\stitle{Very simplified likelihood}
\sdescription{A terse rmarkdown document that  recreates the figures and salient results of the submission}
\end{supplement}

### 2024-03-13 inst/very_simplified_likelihood.Rmd
tidyup
tidyup in prep. for submission

@@

```{r,label=likevis,cache=F}
plotter <- function(...){
a <- seq(from=0,to=1,by=0.01)
n <- 7
jj <- f_vec_arn(a,3,n,log=FALSE)
plot(a,jj/max(jj,na.rm=TRUE),type='n')
grid()
for(r in seq(from=0,to=n)){
  y <- f_vec_a(a,r,n,log=FALSE)
  if(r==0){y[length(y)] <- 1}  
  y <- y/max(y,na.rm=TRUE)
  if(r>0){y[length(y)] <- 0}
  points(a,y,type='l',lwd=4,col=rainbow(n+1)[r+1])
  abline(v=0.64)
}
}
plotter()
pdf(file="ninelikes.pdf")
plotter()
dev.off()
```

@@

Above we see three parkrun results.  In parkrun number 148, I placed
229 out of a field of 318.

```{r optimsinglelike,cache=TRUE}
```

>>>>

\[
\mbox{prob of coming third} =
\frac{b}{a+nb}\cdot
\frac{b}{a+(n-1)b}\cdot
\frac{a}{a+(n-2)b}\cdot
\frac{b}{(n-2)b}\cdots
\frac{b}{2b}\cdot\frac{b}{b}
=\frac{n!}{(n-2)!}\cdot X
\]

Now consider the case where there are $r$ clones coming ahead of the
focal competitor:

\[
\mbox{prob of having $r$ clones ahead}
=\frac{n!}{(n-r)!}\cdot X
\]

@@

```{r,label=likevis,cache=F}
plotter <- function(...){
a <- seq(from=0,to=1,by=0.01)
n <- 8
jj <- f_vec_arn(a,3,n,log=FALSE)
plot(a,jj/max(jj,na.rm=TRUE),type='n',xlab=expression(a),ylab="likelihood")
grid()
rain <- rainbow(n+1)
for(r in seq(from=0,to=n)){
  y <- f_vec_a(a,r,n,log=FALSE)
  if(r==0){y[length(y)] <- 1}
  y <- y/max(y,na.rm=TRUE)
  if(r>0){y[length(y)] <- 0}
  points(a,y,type='l',lwd=4,col=rain[r+1])
}
abline(v=MLE(r=4,n=8))
text(0.07,0.95,"r=8",col=rain[9])
text(0.20,0.95,"r=7",col=rain[8])
text(0.26,0.91,"r=6",col=rain[7])
text(0.29,0.83,"r=5",col=rain[6])
text(0.33,0.76,"r=4",col=rain[5])
text(0.35,0.64,"r=3",col=rain[4])
text(0.39,0.53,"r=2",col=rain[3])
text(0.45,0.41,"r=1",col=rain[2])
text(0.63,0.22,"r=0",col=rain[1])
}

plotter()
pdf(file="ninelikes.pdf")
plotter()
dev.off()
```

```{r shown4n8mle}
MLE(r=4,n=8)
```

@@

Above we see some parkrun results.  In parkrun number 148, I placed
229 out of a field of 318.
